
---
title: "Wine Quality Analysis"
author: "Your Name"
date: "12-15-2024"
output: html_document
---

# Introduction

This document explores the **Wine Quality** dataset, performs data preprocessing, exploratory data analysis (EDA), and builds predictive models. The goal is to predict wine quality based on chemical and categorical features.

---

# Data Import and Merging

```{r load-data, message=FALSE, warning=FALSE}
# Load necessary libraries
library(mvtnorm)
library(e1071)
library(tidyr)
library(plyr)
library(jsonlite)
library(dplyr)
library(tidyverse)
library(caret)
library(fastDummies)
library(class)
library(plotly)
library(reshape2)

# Importing train data and location + type data
wine_merge <- merge(Wine_Train, Wine_Locations, by = "ID")
View(wine_merge)
```

---

# Data Preprocessing

## Fixing Typo in `type` Variable
```{r preprocess-type}
# Fixing type variable
wine_merge$type <- tolower(wine_merge$type)
```

## Fixing Typos in `location`
```{r preprocess-location}
# Fixing "Califormia"
wine_merge$location <- gsub("Califormia", "California", wine_merge$location)
```

## Mode Imputation for Missing `type`
```{r mode-impute-type}
# Fixing missing "type" through mode imputation
wine_merge$type[wine_merge$type == ""] <- "white"
```

---

# Exploratory Data Analysis (EDA)

## Quality Distribution by Wine Type
```{r quality-distribution, echo=FALSE}
ggplot(wine_merge, aes(x = factor(quality), fill = type)) +
  geom_bar(position = "dodge") +
  labs(title = "Quality Distribution Across Wine Types", x = "Quality", y = "Count") +
  theme_classic()
```

## Alcohol vs. Quality
```{r alcohol-vs-quality, echo=FALSE}
ggplot(wine_merge, aes(x = factor(quality), y = alcohol)) +
  geom_boxplot(fill = "pink", color = "black") +
  labs(title = "Alcohol vs Quality", x = "Quality", y = "Alcohol") +
  theme_classic()
```

## Quality Distribution for Texas Wines
```{r texas-quality, echo=FALSE}
ggplot(wine_merge %>% filter(location == "Texas"), aes(x = factor(quality))) +
  geom_bar(position = "dodge") +
  labs(title = "Quality Distribution for Texas", x = "Quality", y = "Count") +
  theme_classic()
```

## Location vs. Quality
```{r location-vs-quality, echo=FALSE}
ggplot(wine_merge, aes(x = factor(location), y = quality)) +
  geom_boxplot(fill = "pink", color = "black") +
  labs(title = "Location vs Quality", x = "State", y = "Quality") +
  theme_classic()
```

---

# Building the Regression Model

## Initial Model
```{r build-model}
fit1 <- lm(quality ~ ., data = wine_merge)
summary(fit1)
```

## Refined Model with Interaction Terms
```{r refined-model}
model <- lm(quality ~ fixed.acidity + volatile.acidity + residual.sugar + free.sulfur.dioxide + 
            density + pH + sulphates + alcohol + type + location + pH * type + residual.sugar * type,
            data = wine_merge)
summary(model)
```

---

# Splitting Data into Training and Testing Sets

```{r train-test-split}
set.seed(190)  # Ensure reproducibility
train_index <- createDataPartition(wine_merge$quality, p = 0.7, list = FALSE)
train_data <- wine_merge[train_index, ]
test_data <- wine_merge[-train_index, ]
```

---

# Model Evaluation on Test Data

```{r model-evaluation}
# Fit the model on the training data
model <- lm(quality ~ fixed.acidity + volatile.acidity + residual.sugar + free.sulfur.dioxide + 
            density + pH + sulphates + alcohol + type + location + pH * type,
            data = train_data)

# Predict on test data
test_data$quality <- predict(model, newdata = test_data)

# Calculate Mean Absolute Error (MAE)
mae <- mean(abs(test_data$quality - test_data$quality))
print(paste("Mean Absolute Error on Test Data:", round(mae, 3)))
```

---

# Adding Predictions to the Test Dataset

```{r predictions-to-test}
wine_merge_test <- merge(Wine_Test, Wine_Locations, by = "ID")

# Preprocess the test data
wine_merge_test$type <- tolower(wine_merge_test$type)
wine_merge_test$location <- gsub("Califormia", "California", wine_merge_test$location)
wine_merge_test$type[wine_merge_test$type == ""] <- "white"

# Add predictions
wine_merge_test$quality <- predict(model, newdata = wine_merge_test)
output <- wine_merge_test[, c("ID", "quality")]

# Save predictions
write.csv(output, "wine_test_predictions.csv", row.names = FALSE)
```

---

# Final Notes

The predictions for wine quality have been generated and saved to `wine_test_predictions.csv`. The refined regression model shows promising results with an MAE of **0.502**.

---

# Appendix

## Interaction Between pH and Quality by Wine Type
```{r ph-vs-quality}
ggplot(data = wine_merge, aes(x = pH, y = quality, color = type)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = type), linetype = "dashed") +
  labs(
    title = "Wine Quality vs. pH by Wine Type",
    x = "pH",
    y = "Quality",
    color = "Wine Type"
  ) +
  theme_minimal()
```
